mixchar

### 2018-08-15 manuscript/manuscript.Rnw
add stt examples

\textbf{\textit{Deconvolution}}\\
\textit{Rate of mass loss}\\
After we've loaded our data, we need to calculate the rate of mass loss by taking the derivative. The \verb|process| function needs the dataset, the initial mass of sample, the name of the temperature data column, and the name of the mass column (mg).

@@

The output of the \verb|process| function is a list of three items: the modified dataframe including the derivative thermogravimetric rate of mass loss data (DTG), the initial mass value that was supplied, and the maximum and minimum temperature of values in the data. Plotting this output will results in two curves: the mass of sample across temperature and the rate of mass loss curve (Fig.~\ref{Fig:rateloss}). The rate of mass loss curve is a multi-peaked curve encompassing three main phases \citep{orfao2001}:

@@

\verb|deconvolve| uses non-linear optimisation with residual sum of squares to fit the rate expression \citetext{as in \citealp{cheng2015}}. Starting values for all parameters were selected based on curves depicted in the literature \citep{muller-hagedorn2007} and from the results of running an identical deconvolution on pure cellulose and lignin samples. Hemicelluloses decay in a reasonably narrow band beginning at a lower temperature \citep{muller-hagedorn2007}, so we used 270 for position and 50 for width. Linear cellulose crystals decay at a higher temperature, but decay more rapidly after peak temperatures are reached, so starting position was set to 310 and width to 30. Lignin typically decays beginning at a high temperature and over a wide interval \citep{chen2015}, so position and width began at 410 and 200, respectively. These starting values can also be modified in the \verb|deconvolve| function with the \verb|start_vec|, \verb|lower_vec|, and \verb|upper_vec| arguments.

@@

The output of the \verb|deconvolve| function is a list of five items:
\begin{enumerate}
\item{the data used to fit the function,
<<>>=
DTG_data <- rate_data(output_juncus)
head(DTG_data)
@
}
\item{the upper and lower temperature bounds to which the data were cropped,
<<>>=
temp_bounds(output_juncus)
@

}
\item{the model fit including parameter estimates,
<<eval = FALSE>>=
model_fit(output_juncus)
@

<<echo = FALSE>>=
cat(paste0('Nonlinear regression model\n',
          '  model: deriv ~ fs_mixture(temp_C, height_1, skew_1, position_1,\n',
          '    width_1, height_2, skew_2, position_2, width_2, height_3,\n',
          '    skew_3, position_3, width_3)\n',
          '  data: dataframe\n',
          '  height_1   skew_1     position_1 width_1    height_2   skew_2\n',
          '  3.944e-03  1.258e-01  2.662e+02  5.106e+01  5.793e-03  1.344e-02\n',
          '  position_2 width_2    height_3   skew_3     position_3 width_3\n',
          '  3.173e+02  2.866e+01  1.163e-03  1.085e-01  3.300e+02  2.500e+02\n',
          '  residual sum-of-squares: 9.299e-06\n\n',
          'Number of iterations to convergence: 23\n',
          'Achieved convergence tolerance: 1.49e-08'))
@
}
\item{the number of peaks,
<<>>=
output_juncus$n_peaks
@
}
\item{and the mean, 2.5\% and 97.5\% estimates, median, and standard deviation of the weight of each component.
<<>>=
component_weights(output_juncus)
@
}
\end{enumerate}
If you \verb|plot()| the output of the \verb|deconvolve| function you will see the result of the deconvolution (Fig.~\ref{Fig:decon}). This default plot shows the underlying DTG data, the overall mixture model curve, as well as the components peaks. The default plot is in black and white, but you can plot a colour version that uses colour-blind friendly \verb|viridis| colours \citep{viridis} by specifying \verb|bw = FALSE|.

@@

The Fraser-Suzuki family of functions are exported to allow users to create their own plots from the model outputs. Detailed descriptions of how to do so are included in the package vignettes.

<<>>=
stt <- read.table("../mixchar.notes/USATtLeaf_data.txt", fill = TRUE, header = FALSE, skip = 46, skipNul = TRUE, fileEncoding="UTF-16LE")
colnames(stt) <- c("time", "temp", "weight", "deriv")
@

@@

1. How robust is this (testing with Stacy’s data)
show with a few samples
- to show different machine
- to show a variety of sample types
how to include these in the manuscript?
Discuss limitations of estimates
how reliable are they?

4. Future work
The package was developed using thermogravimetric data for 29 different wetland plant species, as well as pure samples of cellulose (carboxy-methyl cellulose) and lignin (alkali lignin from Sigma Aldrich). These data came from the same TGA-FTIR machine and were exported in the same format. For this reason, the package was also tested on thermogravimetric data exported from a different machine in a different format, on different species and plant parts in order to test behaviour of the package.

Below we present results from

add in discussion that happy to develop for data of other machines/in other formats

tools that already exist
mixtools
how does my approach compare with traditional mixture model estimation?

Statistical inference of mixture models

\section*{Quality control}
All the functions of \verb|mixchar| were tested to see if they produced the desired output.

@@

For those that wish to contribute to the package, it is hosted on Github. Contributors can log issues via the issues tracker (\url{https://github.com/smwindecker/mixchar/issues}) or submit a pull request to add functionality to the package.

\section*{Acknowledgements}
The authors would like to thank Dr. Nick Tierney and David Wilkinson for reviewing an early version of the package. We would like to thank volunteers Paula Sanchez, Abbey Kinnish, Kelsey Johnson, Madeline Brenker, and Urtzi Enriquez Urzelai for assistance in the field collecting plant specimens. We would like to thank the University of Melbourne Department of Chemical and Biomolecular Engineering for access to and training on the TGA-FTIR. Vegetation collection for data used to build and verify the model was conducted under Victorian Department of Environment, Land, Water and Planning Permit No 10007429.

>>>>

\textbf{\textit{Deconvolution}}\\
\textit{Rate of mass loss}\\
After we've loaded our data, we need to calculate the rate of mass loss by taking the derivative. The \verb|process| function needs the dataset, the initial mass of sample, the name of the temperature data column, and the name of the mass column (mg), which can be specified either as mass loss data with the \verb|mass_loss| argument or as mass data with the \verb|mass| argument.

@@

The output of the \verb|process| function is a list of three items: the modified dataframe including the derivative thermogravimetric rate of mass loss data (DTG), the initial mass value that was supplied, and the maximum and minimum temperature of values in the data. Plotting this output results in two curves: the mass of sample across temperature and the rate of mass loss (Fig.~\ref{Fig:rateloss}). The rate of mass loss is a multi-peaked curve encompassing three main phases \citep{orfao2001}:

@@

\verb|deconvolve| uses non-linear optimisation with residual sum of squares to fit the rate expression \citetext{as in \citealp{cheng2015}}. Starting values for all parameters were selected based on curves depicted in the literature \citep{muller-hagedorn2007} and from the results of running an identical deconvolution on pure cellulose and lignin samples. Hemicelluloses decay in a reasonably narrow band beginning at a lower temperature \citep{muller-hagedorn2007}, so we used 270 for position and 50 for width. Linear cellulose crystals decay at a higher temperature, but decay more rapidly after peak temperatures are reached, so starting position was set to 310 and width to 30. Lignin typically decays beginning at a high temperature and over a wide interval \citep{chen2015}, so position and width began at 410 and 200, respectively. Default starting values were tested on litter from 30 plant species, encompassing herbaceous, graminoid, as well as woody species. However, they can also be modified with the \verb|start_vec|, \verb|lower_vec|, and \verb|upper_vec| arguments if needed.

@@

\textbf{\textit{Package outputs}}\\
The output of the \verb|deconvolve| function is a list of five items that can be retrieved with accessor functions:
\begin{enumerate}
\item{the processed dataset that results from the \verb|process| function, useful if for playing around with other modelling approaches or plotting options, and accessed with \verb|rate_data()|:
<<>>=
DTG_data <- rate_data(output_juncus)
head(DTG_data)
@
}
\item{the temperature values at which the data were cropped for analysis, accessed with \verb|temp_bounds()|:
<<>>=
temp_bounds(output_juncus)
@

}
\item{the output of the mixture model and estimated parameter values for height, skew, position, and width for each peak. Peak 1 is hemicellulose, peak 2 is cellulose, and peak 3 is lignin. If present, the optional fourth peak located at the lowest temperature interval will be listed as peak 0. Accessed with \verb|model_fit()|:
<<eval = FALSE>>=
model_fit(output_juncus)
@

<<echo = FALSE>>=
cat(paste0('Nonlinear regression model\n',
          '  model: deriv ~ fs_mixture(temp_C, height_1, skew_1, position_1,\n',
          '    width_1, height_2, skew_2, position_2, width_2, height_3,\n',
          '    skew_3, position_3, width_3)\n',
          '  data: dataframe\n',
          '  height_1   skew_1     position_1 width_1    height_2   skew_2\n',
          '  3.944e-03  1.258e-01  2.662e+02  5.106e+01  5.793e-03  1.344e-02\n',
          '  position_2 width_2    height_3   skew_3     position_3 width_3\n',
          '  3.173e+02  2.866e+01  1.163e-03  1.085e-01  3.300e+02  2.500e+02\n',
          '  residual sum-of-squares: 9.299e-06\n\n',
          'Number of iterations to convergence: 23\n',
          'Achieved convergence tolerance: 1.49e-08'))
@
}
\item{the number of peaks:
<<>>=
output_juncus$n_peaks
@
}
\item{and the mean, 2.5\% and 97.5\% estimates, median, and standard deviation of the weight of each component that can be accessed with \verb|component_weights()|:
<<>>=
component_weights(output_juncus)
@
}
\end{enumerate}

Plotting the output of the \verb|deconvolve| function shows the underlying DTG data, the overall mixture model curve, as well as the components peaksof the deconvolution (Fig.~\ref{Fig:decon}). The default plot is in black and white, but you can plot a colour version that uses colour-blind friendly \verb|viridis| colours \citep{viridis} by specifying \verb|bw = FALSE|.

@@

The Fraser-Suzuki family of functions are exported to allow users to create their own plots from the model outputs. You can use these in conjunction with the parameter estimates, accessed as follows:

<<>>=
juncus_parameters <- model_parameters(output_juncus)
@

@@

The package was developed using thermogravimetric data for 29 different wetland plant species, as well as pure samples of cellulose (carboxy-methyl cellulose) and lignin (alkali lignin from Sigma Aldrich). These data came from the same TGA-FTIR machine and were exported in the same format. For this reason, the package was also tested on thermogravimetric data exported from a different machine in a different format, on different species and plant parts in order to test behaviour of the package.
1. How robust is this (testing with Stacy’s data)
show with a few samples
- to show different machine
- to show a variety of sample types
how to include these in the manuscript?
Discuss limitations of estimates
how reliable are they?

\section*{Quality control}
Something about passing CRAN check with no errors or warnings, or notes.

All the functions of \verb|mixchar| were tested to see if they produced the desired output. The workflow was tested on thermogravimetric data from two different TGA-FTIR machines, and on samples outside those used to build the package.

\textit{Seagrass blade}
<<echo = FALSE>>=
source('stt_analysis.R')
sg_leaf <- process_data('data/seagrass_leaf.txt')
@

after loading the data, and using the process function, we can inspect the plots:
<<>>=
plot(sg_leaf)
@

we can determine that this sample needs a fourth peak, and the lower and upper bounds modified.
<<>>=
sg_out <- deconvolve(sg_leaf,
                     n_peaks = 4,
                     lower_temp = 150,
                     upper_temp = 600)
plot(sg_out, bw = FALSE)
@
the outcome looks good.

\textit{Seagrass rhizome}
similarly needed 4 peaks specified and bounds.
<<echo = FALSE>>=
sg_rhiz <- process_data('data/seagrass_rhizome.txt')
@

<<>>=
sg_rh_out <- deconvolve(sg_rhiz,
                        n_peaks = 4,
                        lower_temp = 150,
                        upper_temp = 600)
plot(sg_rh_out, bw = FALSE)
@

\textit{Seagrass root}
<<echo = FALSE>>=
sg_root <- process_data('data/seagrass_root.txt')
@

<<>>=
sg_root_out <- deconvolve(sg_root,
                          lower_temp = 150,
                          upper_temp = 600)
plot(sg_root_out, bw = FALSE)
@

\textit{Macroalgae}
<<echo = FALSE>>=
ma_leaf <- process_data('data/macroalgae_leaf.txt')
@

<<>>=
plot(ma_leaf)
@

<<>>=
start_vec <- c(0.002, -0.15, 250, 50,
               0.003, -0.15, 310, 50,
               0.006, -0.15, 350, 30,
               0.001, -0.15, 410, 200)
ub <- c(2, 0.2, 260, 80,
        2, 0.2, 330, 90,
        2, 0.2, 380, 50,
        2, 0.2, 430, 250)
ma_out <- deconvolve(ma_leaf,
                     n_peaks = 4,
                     start_vec = start_vec,
                     upper_vec = ub,
                     lower_temp = 150,
                     upper_temp = 600)
plot(ma_out, bw = FALSE)
@

@@

For those that wish to contribute to the package, it is hosted on Github. Contributors can log issues, for example with alternative data foramts, via the issues tracker (\url{https://github.com/smwindecker/mixchar/issues}) or submit a pull request to add functionality to the package.

\section*{Acknowledgements}
The authors would like to thank Dr. Nick Tierney and David Wilkinson for reviewing an early version of the package as well as CRAN reviewers for useful additions. We would like to thank volunteers Paula Sanchez, Abbey Kinnish, Kelsey Johnson, Madeline Brenker, and Urtzi Enriquez Urzelai for assistance in the field collecting plant specimens. We would like to thank the University of Melbourne Department of Chemical and Biomolecular Engineering for access to and training on the TGA-FTIR. Vegetation collection for data used to build and verify the model was conducted under Victorian Department of Environment, Land, Water and Planning Permit No 10007429.
